Transport map accelerated adaptive importance sampling, and application to inverse problems arising from multiscale stochastic reaction networks

### 2018-06-12 TPAIS_2018.tex
Updating

\author{Simon L. Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e:
    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
    grant award EP/L023393/1. SLC would like to thank the Isaac Newton
    Institute for Mathematical Sciences for support and hospitality
    during the programme ``Uncertainty quantification for complex systems: theory and methodologies'' when work on this paper was undertaken. This work was supported by:
EPSRC grant number EP/K032208/1.} \and Ioannis G. Kevrekidis\thanks{} \and Paul
  Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK.}}

@@

\section{Introduction}
%Topics to cover: parallel MCMC, pMC, PAIS
%Transport maps, transport map MCMC
%Stochastic reaction networks
%Multiscale approximations, QSSA/QEA, CMA

In Section \ref{sec:PAIS}, we will briefly reintroduce Parallel
Adaptive Importance Sampling (PAIS), a variant of Population Monte
Carlo (PMC) which incorporates state of the art resamplers. In Section \ref{sec:map} we show how an appropriate transport map can
be constructed from importance samples which maps the posterior close
to a reference Gaussian measure. In Section \ref{sec:TPAIS} we show
how such a map can be incorporated into a sophisticated parallel MCMC
infrastructure in order to accelerate mixing. In Section
\ref{sec:conv} we seek to show the advantages of this approach through
the analysis of a challenging test problem. In Section
\ref{sec:multi} we consider how likelihoods can be approximated using
 multiscale methodologies in order to carry out inference for
multiscale and/or partially observed stochastic reaction networks. In
Section \ref{sec:num} we present some numerical examples, which serve
to demonstrate the increased efficiency of the described sampling
methodologies, as well as investigating the posterior approximations
discussed in the previous section. We conclude with a discussion in
Section \ref{sec:conc}.

>>>>

\author{Simon L. Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e:
    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
    grant award EP/L023393/1. SLC would like to thank the Isaac Newton
    Institute for Mathematical Sciences for support and hospitality
    during the programme ``Uncertainty quantification for complex systems: theory and methodologies'' when work on this paper was undertaken. This work was supported by:
EPSRC grant number EP/K032208/1.} \and Ioannis
G. Kevrekidis\thanks{Chemical and Biomolecular Engineering \& Applied
  Mathematics and Statistics, Johns Hopkins University, Baltimore, MD
  21218. The work of IGK was partially supported by the US National Science Foundation and by DARPA.} \and Paul
  Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK.}}

@@

\section{Introduction}
The importance of Markov chain Monte Carlo (MCMC) methods is becoming
increasingly apparent, in a world replete with datasets which need to be
combined with complex dynamical systems in order for us to be able
make progress in a range of scientific fields. Different MCMC methods
have been designed with different challenges in mind; high dimensional
targets****CITE FASTER PAPER*****, large data sets ****zigzag****,
other methods which exploit piecewise deterministic Markov processes
***bouncey particle***,
methods which are able to make very large moves informed by the
data****stochastic mala, HMC****, methods which are able to exploit
the geometry of the target***riemannian****. Population Monte Carlo
(pMC) methods uses efficient importance sampling proposals informed by an
ensemble of particles which have already learned about the structure
of the target***cite standard papers including AIS etc****. These methods are particularly useful in the context of
low dimensional target distributions which have complex structure, for
which traditional Metropolis-Hastings methods tend to struggle, for
example when the target is multimodal or if the density is
concentrated on a lower-dimensional manifold. Moreover, they are an
excellent candidate for parallelisation, since the task of computing
the likelihood for each particle in the ensemble can be distributed
across a set of processors. This is of utmost importance in an age
where increases in computational power are most likely to come from
large-scale parallelisation as opposed to the large speed-ups in
individual processors that we saw throughout the 20th Century.

In ****cite PAIS***, we explored ideas from adaptive MCMC within a
pMC-type structure, in order to construct an adaptive importance
sampler which could automatically adapt to optimise
efficiency. Alongside this, we explored the use of state-of-the-art
resamplers****cite sebastian***, and a greedy approximation (the
approximate multinomial resampler) in order to improve the quality of
the importance sampler proposals.

Other approaches to challenging posterior structure have been
proposed, including the work of Marzouk and Parnod***cite***, who
demonstrated how invertible transport maps can be constructed which
map the target distribution close to an easily-explored reference
density, such as a d-dimensional Gaussian. Since the map is
invertible, Gaussian proposal densities on the reference space can be
mapped to highly informed moves on the parameter space.

In this paper, we particularly concern ourselves with the problem of
inverse problems in stochastic reaction networks. This type of model
is used in situations where ODE or diffusion****Chem langevin****
approximations of chemical reaction kinetics are not viable, usually
because of the low concentration of one or more chemical
species****Gillespie****. Biochemical networks typically fall into
this category, because of the small volume of the reactors involved (a
cell).

One challenge with networks arising from these applications is that
they often have some reactions which are occurring many times on a
timescale in which other reactions in the system are unlikely to
occur. This multiscale behaviour throws up two main challenges, both
in observability in an experimental setting, but also in the context
of modelling, where multiscale systems can make the simulation of the
continuous time Markov chain computationally intractable.

A range of numerical methods have been suggested to tackle this latter
problem, including tau-leaping***cite****, diffusion approximations
***radek and Yannis, cotter and Yannis****, quasi-steady state
approximations ****slow scale and nested****, and other averaging
arguments ****cotter comp. phys****. Some of these approaches are more
computationally intensive than others, but this can be mitigated with the
use of analytical results about the invariant densities of certain
types of fast subsystems***product-form papers****.

We would not expect the fast processes in such a system to be
experimentally observable. In such a situation where only the slow
variables in a system are identifiable, in an ideal world, we would be
be able to integrate over the space of all possible trajectories of
the fast processes. Since this is an insurmountable computational
task, we might instead wish to approximate the likelihoods arising
from such a set of observations by the use of an appropriate
multiscale methodology.

The partial observability of these systems leads to only
partial-observability of the unknown parameters in the system, for
example the reaction rates of the fast reactions. The resulting
multiscale approximations, and the resulting approximations of the
likelihoods arising from the inverse problem for such a network, then
often exhibit 
insensitivity to changes in certain directions of
parameter space. This causes problems when aiming to sample from such
a target, since the density is concentrated on a manifold whose
tangential vectors point in directions of insensitivity. This
\emph{sloppiness}***cite*** of the likelihood is a challenging problem
that appears in many applications in science and engineering***active
subspaces literature here****.

In this work, we aim to improve the PAIS machinery, which is limited,
as the same proposal kernel is used for each particle in the
ensemble. This homogeniety of proposal kernels means that complex
structures in the density can only be well represented with a large
increase in the ensemble size, increasing computational cost. If
computational resources do not allow for such an increase, this can
lead to stability issues in the algorithm, due to the proposal
distribution in the importance sampler not representing well the
target density, leading to high variance in the weights, causing slow
convergence.

As shown in ***cite marzouk****, transport maps can be used to create
proposal densities which are highly informed by the structure of the
target density. In this paper, we demonstrate how such maps can also
be constructed using importance weights arising from a pMC-type
algorithm in order to accelerate and stabilise convergence. We will
then apply the methodology to some complex posterior distributions
arising from Bayesian inversion of multiscale (bio-)chemical networks.

%Topics to cover: parallel MCMC, pMC, PAIS
%Transport maps, transport map MCMC
%Stochastic reaction networks
%Multiscale approximations, QSSA/QEA, CMA

In Section \ref{sec:PAIS}, we will briefly reintroduce PAIS. In Section \ref{sec:map} we show how an appropriate transport map can
be constructed from importance samples which maps the posterior close
to a reference Gaussian measure. In Section \ref{sec:TPAIS} we show
how such a map can be incorporated into a sophisticated parallel MCMC
infrastructure in order to accelerate mixing, and improve stability. In Section
\ref{sec:conv} we seek to show the advantages of this approach through
the analysis of a challenging test problem. In Section
\ref{sec:multi} we consider how likelihoods can be approximated using
 multiscale methodologies in order to carry out inference for
multiscale and/or partially observed stochastic reaction networks. In
Section \ref{sec:num} we present some numerical examples, which serve
to demonstrate the increased efficiency of the described sampling
methodologies, as well as investigating the posterior approximations
discussed in the previous section. We conclude with a discussion in
Section \ref{sec:conc}.
